% Claim proof file for row claim_3_13 -- "AcyclicityUnderIntervention".
% Auto-generated stub by scaffold/solving_current_row.py at row start. A
% manager agent dedicated to writing the TeX proof fills in the body below.
% The proof must:
%   - Stay close to the lecture notes' proof if one exists (always search
%     the LN first for a `\begin{proof}` block immediately following the
%     claim; copy / lightly edit when present).
%   - Otherwise use the LN paradigm and cite prior claims by their `ref`.
%   - Be self-contained enough that a mathematician can follow it without
%     re-reading the LN.
%   - Have no `sorry`, `omitted`, or "obvious" shortcuts.
%
% Compiles standalone via the `subfiles` package.

\documentclass[main]{subfiles}

\begin{document}
% ref: claim_3_13 (proof, with statement restated for readability)
% title: AcyclicityUnderIntervention
\def\rowref{claim\_3\_13}
\def\rowtitle{AcyclicityUnderIntervention}
\phantomsection\label{claim_3_13}
%
% Cross-references: see claim_<ref>_statement_<title>.tex in this folder
% for the corresponding statement.

% --- statement (restated) ---------------------------------------------------
\begin{Rem}
    If a CDMG $G=(J,V,E,L)$ is acyclic then also $G_{\doit(I_W)}$ is acyclic and a topological order for
    $G_{\doit(I_W)}$ is also one for $G$.
    Any topological order of $G$ can be extended to one for $G_{\doit(I_W)}$, e.g.\
    by putting all the $I_w$ nodes first in the ordering.
\end{Rem}

% --- proof ------------------------------------------------------------------
% Proof constructed from scratch in the LN paradigm: the LN's \Rem at
% lecture-notes/lecture_notes/graphs.tex:821--829 contains no \begin{proof}
% block (the next claimmark, claim_3_14, follows immediately at line 831).
% The three parts mirror the three theorems in the companion Lean file
% AcyclicityUnderInterventionNodes.lean (parts A / B / C correspond to
% isAcyclic_..., ..._restrict, ..._extend respectively); the named order in
% Part C mirrors \refrow{claim_3_6}'s standalone splitOrder relation.
\begin{proof}
The whole argument rests on a single edge-set observation: by \refrow{def_3_13} item iii,
\[
  E_{\doit(I_W)}
  \;=\;
  E \;\dcup\; \lC I_w \tuh w \,\middle|\, w \in W \sm J \rC,
\]
i.e.\ every edge of $G_{\doit(I_W)}$ is either an original edge of $G$ or a fresh edge $I_w \tuh w$ for some $w \in W \sm J$. In particular, the \refrow{def_3_1} invariant $E \ins (J \cup V) \times V$ together with the fresh-edge target $w \in W \sm J \ins V$ (using the precondition $W \ins J \cup V$ to rule out $w \in W \sm J$ being outside $G$, leaving $w \in V$) gives that \emph{every} edge of $G_{\doit(I_W)}$ has its target in $V$:
\[
  E_{\doit(I_W)} \;\ins\; \lp J_{\doit(I_W)} \cup V_{\doit(I_W)} \rp \times V.
\]
By contrast, the \emph{source} of an edge can be either an original node $v' \in J \cup V$ (first disjunct, $E$-piece) or a fresh intervention node $I_w$ (second disjunct). This source--target asymmetry is the key observation driving all three parts below.

\medskip\noindent\emph{Part A (acyclicity).}
Suppose, for contradiction, that $G_{\doit(I_W)}$ is not acyclic. By \refrow{def_3_6} there is a node $v$ of $G_{\doit(I_W)}$ admitting a non-trivial directed walk to itself:
\[
  v = v_0 \tuh v_1 \tuh \cdots \tuh v_n = v
  \qquad \text{in } G_{\doit(I_W)}, \quad n \ge 1,
\]
i.e.\ $(v_k, v_{k+1}) \in E_{\doit(I_W)}$ for every $k = 0, \dots, n-1$ (\refrow{def_3_4}).

In a directed cycle every vertex is the target of some edge: for each $k \in \lC 1, \dots, n \rC$ the step $(v_{k-1}, v_k) \in E_{\doit(I_W)}$ has $v_k$ as its target, and the closure $v_0 = v_n$ then makes $v_0 = v_n$ a target as well. By the edge-set observation above, every target of an edge of $G_{\doit(I_W)}$ lies in $V$. Hence $v_k \in V$ for all $k = 0, \dots, n$; in particular no $v_k$ is one of the fresh $I_w$ intervention nodes.

We now show each step is an original $E$-edge. Fix $k$ and consider $(v_{k-1}, v_k) \in E_{\doit(I_W)}$. By \refrow{def_3_13} item iii it is either an original edge $(v_{k-1}, v_k) \in E$, or a fresh edge $I_w \tuh w$ for some $w \in W \sm J$. The fresh-edge case is impossible: its source is $I_w$, a fresh intervention node, but we just showed $v_{k-1} \in V$ is original. So $(v_{k-1}, v_k) \in E$.

The same vertex sequence
\[
  v = v_0 \tuh v_1 \tuh \cdots \tuh v_n = v
  \qquad \text{in } G
\]
is therefore a non-trivial directed walk in $G$ (\refrow{def_3_4}). Since $v_0 \in V$, in particular $v_0$ is a node of $G$ (\refrow{def_3_2}), so $v_0$ admits a non-trivial directed walk to itself in $G$, contradicting acyclicity of $G$ (\refrow{def_3_6}).

\medskip\noindent\emph{Part B (restriction direction).}
Assume $W \ins J \cup V$ and that $<'$ is a topological order of $G_{\doit(I_W)}$ (\refrow{def_3_8}). We claim that the restriction
\[
  v < w \;:\iff\; v <' w \qquad (v, w \in J \cup V)
\]
is a topological order of $G$. Note that the canonical embedding $J \cup V \hookrightarrow J_{\doit(I_W)} \cup V_{\doit(I_W)}$ is the identity: by \refrow{def_3_13} items i.--ii., $J_{\doit(I_W)} = J \dcup \lC I_w \mid w \in W \sm J \rC$ and $V_{\doit(I_W)} = V$, so $J \ins J_{\doit(I_W)}$ and $V = V_{\doit(I_W)}$, giving $J \cup V \ins J_{\doit(I_W)} \cup V_{\doit(I_W)}$. Hence every $v \in G$ is also a node of $G_{\doit(I_W)}$ (\refrow{def_3_2}). We verify the four conditions of \refrow{def_3_8} in turn.

\smallskip
\noindent\emph{Irreflexivity on $G$.} For $v \in G = J \cup V$, the embedding gives $v \in G_{\doit(I_W)}$, so $\neg(v <' v)$ by irreflexivity of $<'$. By definition of $<$, this is $\neg(v < v)$. \checkmark

\smallskip
\noindent\emph{Transitivity on $G$.} For $v, w, x \in G$, all three lie in $G_{\doit(I_W)}$ by the embedding. If $v < w$ and $w < x$, unfolding the definition gives $v <' w$ and $w <' x$, hence $v <' x$ by transitivity of $<'$, i.e.\ $v < x$. \checkmark

\smallskip
\noindent\emph{Trichotomy on $G$.} For $v, w \in G$, the embedding gives $v, w \in G_{\doit(I_W)}$, and trichotomy of $<'$ yields $v <' w \lor v = w \lor w <' v$, equivalently $v < w \lor v = w \lor w < v$. \checkmark

\smallskip
\noindent\emph{Parents precede children.} Suppose $v \in \Pa^G(w)$. By \refrow{def_3_5} this means $v \in G$ together with $(v, w) \in E$. The first disjunct of \refrow{def_3_13} item iii embeds $E$ into $E_{\doit(I_W)}$, so $(v, w) \in E_{\doit(I_W)}$. Combined with $v \in G_{\doit(I_W)}$ (from the embedding), \refrow{def_3_5} gives $v \in \Pa^{G_{\doit(I_W)}}(w)$. The topological-order condition for $<'$ on $G_{\doit(I_W)}$ then yields $v <' w$, i.e.\ $v < w$. \checkmark

Combining the four checks, $<$ is a topological order of $G$ in the sense of \refrow{def_3_8}.

\medskip\noindent\emph{Part C (extension direction with explicit recipe).}
Conversely, assume $W \ins J \cup V$ and that $<$ is a topological order of $G$. We construct a topological order $<'$ of $G_{\doit(I_W)}$ by the LN's recipe: put all $I_w$ nodes first, ordered among themselves by the underlying $w \in W \sm J$ values, and inherit $<$ on the original nodes. Explicitly, on the carrier
\[
  J_{\doit(I_W)} \cup V_{\doit(I_W)}
  \;=\; (J \cup V) \,\dcup\, \lC I_w \,\middle|\, w \in W \sm J \rC
\]
(by \refrow{def_3_13} items i.--ii.\ and the precondition $W \ins J \cup V$), define
\[
  a <' b \;:\iff\;
  \begin{cases}
    a < b & \text{if } a, b \in J \cup V \text{ (both original),} \\
    w_1 < w_2 & \text{if } a = I_{w_1}, \ b = I_{w_2} \text{ (both fresh),} \\
    \mathrm{true} & \text{if } a = I_w \text{ (fresh), } b \in J \cup V \text{ (original),} \\
    \mathrm{false} & \text{if } a \in J \cup V \text{ (original), } b = I_w \text{ (fresh).}
  \end{cases}
\]
The four cases are mutually exclusive and exhaustive because $J \cup V$ and $\lC I_w \mid w \in W \sm J \rC$ are disjoint (the second component of the disjoint union in $J_{\doit(I_W)}$, \refrow{def_3_13} item i.). We verify the four conditions of \refrow{def_3_8} by case analysis on which quadrant each vertex lies in.

\smallskip
\noindent\emph{Irreflexivity.} Let $a \in J_{\doit(I_W)} \cup V_{\doit(I_W)}$.
\begin{itemize}
  \item $a \in J \cup V$ (original): $a <' a$ unfolds to $a < a$, false by irreflexivity of $<$ on $G$.
  \item $a = I_w$ (fresh): $a <' a$ unfolds to $w < w$, false by irreflexivity of $<$ on $G$ (using $w \in W \sm J \ins V \ins G$ from the precondition $W \ins J \cup V$).
\end{itemize}
In neither case does $a <' a$ hold. \checkmark

\smallskip
\noindent\emph{Transitivity.} Let $a, b, c \in J_{\doit(I_W)} \cup V_{\doit(I_W)}$ with $a <' b$ and $b <' c$; we show $a <' c$. We label each vertex by its quadrant -- $O$ for original ($\in J \cup V$) or $F$ for fresh ($= I_w$ for some $w \in W \sm J$) -- and case-split on the triple $(a, b, c) \in \lC O, F \rC^3$. The key constraint is that any cross step $O \to F$ has truth value $\mathrm{false}$ (by definition of $<'$), so any case in which either premise $a <' b$ or $b <' c$ is of the form $O \to F$ is vacuous. The eight cases split as follows.
\begin{itemize}
  \item $OOO$: $a <' b$ and $b <' c$ unfold to $a < b$ and $b < c$ in $G$; transitivity of $<$ gives $a < c$, i.e.\ $a <' c$. \checkmark
  \item $FFF$: write $a = I_u, b = I_v, c = I_w$ with $u, v, w \in W \sm J \ins G$ (using the precondition $W \ins J \cup V$). The premises unfold to $u < v$ and $v < w$ in $G$; transitivity of $<$ gives $u < w$, i.e.\ $a <' c$. \checkmark
  \item $FFO$ ($a = I_u, b = I_v, c \in J \cup V$): the first premise unfolds to $u < v$; the second premise is the cross slot $F \to O = \mathrm{true}$. The conclusion $a <' c$ is again the cross slot $F \to O = \mathrm{true}$. \checkmark
  \item $FOO$ ($a = I_u, b, c \in J \cup V$): the first premise is the cross slot $F \to O = \mathrm{true}$; the second premise unfolds to $b < c$. The conclusion $a <' c$ is again the cross slot $F \to O = \mathrm{true}$. \checkmark
  \item $OOF, OFO, OFF, FOF$: each of these has at least one premise of the form $O \to F$, which is $\mathrm{false}$. The hypothesis is vacuous and the implication holds trivially. \checkmark
\end{itemize}
In every non-vacuous case the conclusion $a <' c$ holds. \checkmark

\smallskip
\noindent\emph{Trichotomy.} Let $a, b \in J_{\doit(I_W)} \cup V_{\doit(I_W)}$.
\begin{itemize}
  \item Both original ($a, b \in J \cup V$): trichotomy of $<$ on $G$ gives $a < b \lor a = b \lor b < a$, which unfolds to $a <' b \lor a = b \lor b <' a$.
  \item Both fresh ($a = I_{w_1}, b = I_{w_2}$): trichotomy of $<$ on $G$ at $w_1, w_2 \in W \sm J \ins G$ gives $w_1 < w_2 \lor w_1 = w_2 \lor w_2 < w_1$. The middle disjunct $w_1 = w_2$ gives $I_{w_1} = I_{w_2}$, and the other two unfold to $a <' b$ and $b <' a$.
  \item Cross ($a = I_w$ fresh, $b \in J \cup V$ original): the cross slot $a <' b$ is $\mathrm{true}$, so the first disjunct of trichotomy holds.
  \item Cross ($a \in J \cup V$ original, $b = I_w$ fresh): symmetrically, $b <' a$ is $\mathrm{true}$, so the third disjunct holds.
\end{itemize}
In every case at least one of $a <' b$, $a = b$, $b <' a$ holds. \checkmark

\smallskip
\noindent\emph{Parents precede children.} Suppose $a \in \Pa^{G_{\doit(I_W)}}(b)$, i.e.\ $a \in G_{\doit(I_W)}$ and $(a, b) \in E_{\doit(I_W)}$ (\refrow{def_3_5}). By \refrow{def_3_13} item iii, either $(a, b) \in E$ (an original edge) or $(a, b) = (I_w, w)$ for some $w \in W \sm J$ (a fresh edge).
\begin{itemize}
  \item Original edge $(a, b) \in E$: both endpoints lie in $J \cup V$ (\refrow{def_3_1}), and the source observation $a \in G$ (\refrow{def_3_5}) together with the edge $(a, b) \in E$ yields $a \in \Pa^G(b)$. The topological-order condition for $<$ on $G$ gives $a < b$, which by definition is $a <' b$.
  \item Fresh edge $(a, b) = (I_w, w)$: the source is fresh and the target $w \in W \sm J \ins V \ins J \cup V$ is original. So $a <' b$ unfolds to the cross slot $\mathrm{true}$ and holds directly.
\end{itemize}
In both cases $a <' b$. \checkmark

Combining the four checks, $<'$ is a topological order of $G_{\doit(I_W)}$ in the sense of \refrow{def_3_8}, and by construction it extends $<$ in the sense that the restriction $a <' b \iff a < b$ holds for all $a, b \in J \cup V$ (first quadrant of the case split).
\end{proof}

\end{document}
