\subsection{Do-Calculus}
\totalscore{9}

\begin{figure}[ht]
\centering
\begin{tikzpicture}[scale=1, transform shape]
    \node[ndout] (X) at (0,0) {$X$};
    \node[ndout] (Y) at (4,0) {$Y$};
    \node[ndout] (Z) at (2,2) {$Z$};
    \node[ndout] (W) at (-1,3) {$W$};
    \node[ndout] (L) at (3,4) {$L$};
    \node[ndout] (S) at (0,-2) {$S$};
    \node[ndout] (R) at (4,-2) {$R$};
    \draw[arout] (X) to (Y);
    \draw[arout] (X) to (S);
    \draw[arout] (Y) to (R);
    \draw[arout] (Z) to (X);
    \draw[arout] (Z) to (Y);
    \draw[arout] (W) to (Z);
    \draw[arout] (W) to (X);
    \draw[arout] (L) to (W);
    \draw[arlat, bend right] (W) to (S);
    \draw[arlat, bend left] (L) to (Y);
    \draw[arlat, bend left] (S) to (R);
\end{tikzpicture}
\caption{Acyclic Directed Mixed Graph (ADMG).}
\label{fig:cadmg-1}
\end{figure}

\noindent Consider the observational graph of a causal Bayesian network with latent variables in Figure \ref{fig:cadmg-1}.
%We assume that the variables $X,Y,Z,W,L,S,R$ are observable with probability distribution $P(X,Y,Z,W,L,S,R)$.
To avoid measure-theoretic subtleties, we assume that all occuring Markov kernels are discrete and have strictly positive mass functions.

In this exercise we are interested in estimating the causal effect of $X$ and $W$ on $Y$, that is, the interventional distribution $P(Y|\doit(X,W))$. However, we are only able to perform certain interventional experiments:
\begin{enumerate}[label=(\arabic*)]
    \item We can perform experiments where we intervene on $W$, but this comes, unfortunately, with certain selection bias, $S=s$. In other words, we are only able to produce samples from the conditional interventional distribution: $P(X,Y,Z,L,R|\doit(W),S=s)$ for a fixed value $s$.
    \item We can perform additional unbiased experiments where we intervene on $W$, but there we can only measure the output of $Z$. This means that we can also produce samples from the (unbiased, unconditional)  interventional distribution $P(Z|\doit(W))$.
\end{enumerate}

Use the 3 rules of do-calculus together with the rules of probability (e.g.\ chain rule) to derive an expression for the Markov kernel $P(Y|\doit(X,W))$ in terms of $P(X,Y,Z,L,R|S=s,\doit(W))$ and $P(Z|\doit(W))$. This expression can contain combinations of marginals, conditionals, products and compositions of the latter two Markov kernels. 
Provide all details of your arguments.
When you use d-separation statements clearly indicate in which graph this should hold and draw the graph if it deviates from the one depicted in Figure \ref{fig:cadmg-1}.
%The claimed d-separation statements itself need to be proven as well, by clearly indicating which paths need to be considered and why and where they are d-blocked.
The claimed d-separation statements do not need to be proven (but should be correct, of course).

\emph{Hint: There exists a solution, where the 3 rules of the do-calculus rules are each applied once. Maybe focus on the chain rule of the form $P(Y|\doit(X,W)) = P(Y|U,\doit(X,W)) \circ P(U|\doit(X,W))$ for suitable variables $U$ first and then remove/introduce expressions using the do-calculus rules.}

\answer{%
Since all occuring Markov kernels contain $\doit(W)$ we can restrict to the following CADMG:
\points{1pt for drawing the correct interventional graph with intervention node in the following.}

\centerline{\begin{tikzpicture}[scale=1, transform shape]
    \node[ndout] (X) at (0,0) {$X$};
    \node[ndint] (IX) at (-2,1) {$I_X$};
    \node[ndout] (Y) at (4,0) {$Y$};
    \node[ndout] (Z) at (2,2) {$Z$};
    \node[ndint] (W) at (-1,3) {$W$};
    \node[ndout] (L) at (3,4) {$L$};
    \node[ndout] (S) at (0,-2) {$S$};
    \node[ndout] (R) at (4,-2) {$R$};
    \draw[arout] (X) to (Y);
    \draw[arout] (X) to (S);
    \draw[arout] (Y) to (R);
    \draw[arout] (Z) to (X);
    \draw[arout] (Z) to (Y);
    \draw[arint] (W) to (Z);
    \draw[arint] (W) to (X);
    \draw[arint] (IX) to (X);
    \draw[arlat, bend left] (L) to (Y);
    \draw[arlat, bend left] (S) to (R);
\end{tikzpicture}
}

\begin{enumerate}
    \item By the chain rule we get:
        \begin{align*}
            P(Y|\doit(X,W)) &= P(Y|Z,\doit(X,W)) \circ P(Z|\doit(X,W)).
        \end{align*}
        \points{1pt for the chain rule. TODO: perhaps this should be given as a hint!}
          \item To apply the 3rd rule of do-calculus to $P(Z|\doit(X,W))$ we check the d-separation in the depicted graph $G_{\doit(I_X,W)}$:
         \begin{align*}
             Z \Perp^d_{G_{\doit(I_X,W)}} I_X \given W.
         \end{align*}
         For this one needs to consider the following paths:
         \begin{align*}
             &Z \hut W          && \text{blocked at endpoint $W$,} \\
             &Z \tuh X \hut W   && \text{blocked at endpoint $W$,} \\
             &Z \hut W \tuh X \hut I_X   && \text{blocked at non-collider $W$,} \\
             &Z \tuh X \hut I_X && \text{blocked at collider $X \notin \Anc(W)$,}\\
             &Z \tuh Y \hut X \hut I_X && \text{blocked at collider $Y \notin \Anc(W)$,}\\
             &Z \tuh Y \tuh R \huh S \hut X \hut I_X && \text{blocked at collider $R \notin \Anc(W)$,}
         \end{align*}
         The 3rd rule of do-calculus then shows the equation:
         \begin{align*}
             P(Z|\doit(X,W)) & = P(Z|\doit(W)).
         \end{align*}
         \points{1pt for correct d-separation statement. 1pt for correct application of do-calculus rule.
         %2pt for d-blocked paths (0.5pt for paths involving $W$ are blocked, 0.5pt for every other path (3x)).
         }
     \item To apply the 2nd rule of do-calculus to $P(Y|Z,\doit(X,W))$ we check the d-separation in the depicted graph $G_{\doit(I_X,W)}$: 
         \begin{align*}
             Y \Perp^d_{G_{\doit(I_X,W)}} I_X \given Z,X,W.
         \end{align*}
         First, all paths from $Y$ to $W$ are blocked by $W$ at the endpoint $W$.
         Furthermore, we need to consider the following paths:
         \begin{align*}
             &Y \hut Z \hut W \tuh X \hut I_X && \text{blocked at non-collider $W$,}\\
             &Y \hut Z \tuh X \hut I_X && \text{blocked at non-collider $Z$,}\\
             &Y \hut X \hut I_X && \text{blocked at non-collider $X$,}\\
             &Y \tuh R \huh S \hut X \hut I_X && \text{blocked at non-collider $X$,}
         \end{align*}
         The 2nd rule of do-calculus then shows the equation:
         \begin{align*}
             P(Y|Z,\doit(X,W)) & = P(Y|Z,X,\doit(W)).
         \end{align*}
         \points{1pt for correct d-separation statement.
         1pt for correct application of do-calculus rule.
         %2pt for d-blocked paths (0.5pt for paths involving $W$ are blocked, 0.5pt for every other path (3x)).
         }
     \item To apply the 1st rule of do-calculus to $P(Y|Z,X,\doit(W))$ we check the d-separation in the depicted graph $G_{\doit(I_X,W)}$ (or $G_{\doit(W)}$) : 
         \begin{align*}
             Y \Perp^d_{G_{\doit(I_X,W)}} S \given Z,X,W.
         \end{align*}
        First, all paths from $Y$ to $W$ are blocked by $W$ at the endpoint $W$.
         Furthermore, we need to consider the following paths:
         \begin{align*}
             &Y \hut Z \hut W \tuh X \hut I_X && \text{blocked at non-collider $W$,}\\
             &Y \hut Z \tuh X \hut I_X && \text{blocked at non-collider $Z$,}\\
             &Y \hut X \hut I_X && \text{blocked at non-collider $X$,}\\
             &Y \tuh R \huh S \hut X \hut I_X && \text{blocked at non-collider $X$,}\\
             &Y \hut Z \hut W \tuh X \tuh S && \text{blocked at non-collider $W$,}\\
             &Y \hut Z \tuh X \tuh S && \text{blocked at non-collider $Z$,}\\
             &Y \hut X \tuh S && \text{blocked at non-collider $X$,}\\
             &Y \tuh R \huh S && \text{blocked at collider $R \notin \Anc(Z,X,W)$,}
         \end{align*}
         The 1st rule of do-calculus then shows the equation:
         \begin{align*}
             P(Y|Z,X,\doit(W)) & = P(Y|Z,X,S,\doit(W)).
         \end{align*}
         Since the left hand side is not dependent on $S$ we can evaluate at $S=s$ and get:
         \begin{align*}
             P(Y|Z,X,\doit(W)) & = P(Y|Z,X,S=s,\doit(W)).
         \end{align*}
         \points{1pt for correct d-separation statement. 
           1pt for correct application of do-calculus rule.  
           %3pt for d-blocked paths (1pt for either pointing out paths involving $I_X$ are blocked as before or drawing new graph without $I_X$, 0.5pt for paths involving $W$ are blocked,  0.5pt for every other path (3x)).
         }
     \item Putting all points together we get:
        \begin{align*}
            P(Y|\doit(X,W)) &= P(Y|Z,\doit(X,W)) \circ P(Z|\doit(X,W)) \\
                            &= P(Y|Z,X,S=s,\doit(W)) \circ P(Z|\doit(W)).
        \end{align*}
        \points{1pt for putting it all together.}
\end{enumerate}
}
