\subsection*{ILOs - Experimental Causal Discovery}

We want the students to be able to:
\begin{itemize}
\item explain why correlation does not imply causation.
\item explain ~6 correlation inducing scenarios (e.g.\ cause-effect, effect-cause, feedback, 
    confounding, selection bias, deterministic constraints, etc.).
\item write down and explain the definition of causal effect (in real world terms).
\item explain how one can detect causal relationships experimentally.
\item explain randomized controlled trials and argue about advantages and disadvantages.
\item explain the difference between (conditional) observational distributions and interventional distributions, when they are different, when they are the same.
\item analyse real world studies and claims according to the concepts introduced above.

\end{itemize}


\subsection*{ILOs - Probability and Measure Theory}

We want the students to be able to:
\begin{itemize}
\item work with (elementary) probability theory, write down the main definitions and compute probabilities and expectation values.
\item translate measure theoretic notations, like measure integrals or expectation values, into familiar ones 
    in the discrete and absolute continuous cases.
\item understand, memorize and explain the 'core' definitions and constructions.
\item work with the Euclidean space $\R^D$ (as 'the' standard measurable space).
\item translate measure integrals to the two basic cases of either 1. discrete spaces/random variables/distributions (with probability mass functions) or 2. absolute continuous random variables (with probability densities w.r.t.\ Lebesgue measure) and compute those integrals (with the usual rules of anti-derivatives, etc.).
\item understand and apply the transformation rules for discrete and absolute continuous variables to compute mass functions/densities of push-forward measures (in the absolute continuous case only for differentiable bijections).
\end{itemize}


\subsection*{ILOs - Transition Probability Theory and Markov Kernels}

We want the students to be able to:
\begin{itemize}
\item explain what a Markov kernel is and how to represent them in the discrete and absolute continuous case.
\item give examples of Markov kernels.
\item properly work with discrete Markov kernels and linear Gaussians ones.
\item explain how Markov kernels relate to deterministic maps and probability distributions.
\item explain transition probability spaces and null-sets.
\item explain and work with the 5 constructions of Markov kernels from others 
    (composition, product, marginal, conditional, push-forward) 
    and how to work with them in the discrete and absolute continuous case.
\end{itemize}


\subsection*{ILOs - Conditional Independence}

We want the students to be able to:
\begin{itemize}
\item Write down the definition of conditional independence for conditional random variables 
    and work with them in the discrete and continuous cases.
\item Write down the separoid axioms for conditional independence of conditional random variables.
\item prove separoid axioms for random variables in the discrete and continuous case (in $\R^D$).
\item explain how one can get a given Markov kernel by a deterministic mapping and a probability distribution.
\item construct observational and interventional Markov kernels for SCMs.
\end{itemize}


\subsection*{ILOs - Graph Theory}

We want the students to be able to:
\begin{itemize}
\item write down the definition of CDMGs and explain the meaning.
\item explain all the different concepts of walks (directed, bidirected, collider, bifurcation, path, etc.).
\item explain all the different node/edge relationships (parents, children, etc.)
\item explain the notion of acyclicity
\item write down the definition of topological order and how it relates to acyclicity.
\item explain the different operations on graphs (hard/soft interventions, marginalization)
\item write down the definition of $id$-separation and check if sets are $id$-separated in given graphs.
\item write down the most important separoid axioms for $id$-separation, reason with them and prove them.
\end{itemize}


\subsection*{ILOs - Causal Bayesian Networks}

We want the students to be able to:
\begin{itemize}
\item write down the definition of causal Bayesian network with input and latent variables, motivate the meaning and reason about them.
\item work with the operations on causal Bayesian networks, like hard interventions, soft interventions, marginalization.
\item precisely state the global Markov property, write it down, reason about it and apply it to different causal Bayesian networks where the graph is given.
\item know the main ingredients and steps for the proof of the global Markov property (not necessarily every single step).
\item write down what interventional equivalence means and reason about it.
\item write down how a standard forms (SCM form, canonical form) of causal Bayesian network looks like and reason about it.
\end{itemize}


\subsection*{ILOs - Do-Calculus and Adjustment Formulas}

The student should be able to:
\begin{itemize}
\item write down the 3 do-calculus rules and reason with them.
\item prove the 3 rules of do-calculus using the global Markov property.
\item apply the 3 rules of do-calculus in an iterative manner on (small, not too complicated) graphs for the identification of causal effects (i.e. reducing the number $\doit$-terms).
\item state and apply the back-door covariate adjustment criterion and formula.
\item explain the general idea of the ID-algorithm.
\end{itemize}

\input{ilo-joris.tex}
