\subsection{Simpson's Paradox}
\totalscore{24}
You are investigating the effectiveness of an experimental drug against a new deadly disease.
You are given access to data collected by the RIVM (National Institute for Public Health and the Environment). 
The data concern ``field observations'', so they were not collected in a randomized controlled trial.
You divide the diseased patients into two groups: those that took the drug, and those that didn't take the drug.
Some of the patients recovered, others unfortunately didn't recover and died. 
The reasons why some patients were treated and others were not, are unknown to you.
You find the following numbers:\footnote{Luckily, these numbers are made up.}
\begin{center}\begin{tabular}{l|rr|r|r}
          & Recovery & No recovery & Total & Recovery rate \\
  \hline
  No drug & 1600       & 2400          & 4000    & 40\%       \\
  Drug    & 2000       & 2000          & 4000    & 50\%       \\
  \hline
  Total   & 3600       & 4400          & 8000    &               \\
\end{tabular}\end{center}
Upon closer inspection of the data, you notice something peculiar when you group patients according to sex 
(which we will for simplicity assume to be male or female).
\begin{center}\begin{tabular}{l|rr|r|r}
  Males & Recovery & No recovery & Total & Recovery rate \\
  \hline
  No drug        & 700        & 300           & 1000    & 70\%       \\
  Drug           & 1800       & 1200          & 3000    & 60\%       \\
  \hline
  Total          & 2500       & 1500          & 4000    &               \\
\end{tabular}

\medskip
\begin{tabular}{l|rr|r|r}
  Females & Recovery & No recovery & Total & Recovery rate \\
  \hline
  No drug          & 900        & 2100          & 3000    & 30\%       \\
  Drug             & 200        & 800           & 1000    & 20\%       \\
  \hline
  Total            & 1100       & 2900          & 4000    &               \\
\end{tabular}\end{center}

Interestingly, the probability that a male patient recovers is lower for the treated male patients than for the untreated male patients, and we observe a similar pattern for the female patients.
Yet, if we look at all patients (not conditioning on sex), then the probability to recover is higher for the treated patients than for the untreated.
This phenomenon is an example of \emph{Simpson's paradox}. It stems from our tendency to interpret the observed correlations as causation.

\begin{figure}\centering
\begin{tikzpicture}
  \begin{scope}
    \node[ndout] (X) at (0,0) {$X$};
    \node[ndout] (Y) at (2,0) {$Y$};
    \node[ndout] (Z) at (1,1) {$Z$};
    \draw[tuh] (X) to (Y);
    \draw[tuh] (Z) to (X);
    \draw[tuh] (Z) to (Y);
    \node at (1,-0.75) {(a)};
  \end{scope}
%  \node at (11,1) {\begin{tabular}{ll}
%  $X$ & treatment \\
%  $Z$ & sex \\
%  $Y$ & outcome \\
%  $L$ & unobserved confounder
%  \end{tabular}};
  \begin{scope}[xshift=4cm]
    \node[ndout] (X) at (0,0) {$X$};
    \node[ndout] (Y) at (2,0) {$Y$};
    \node[ndout] (Z) at (1,1) {$Z$};
%    \node[ndlat] (L) at (0,2) {$L$};
    \draw[tuh] (X) to (Y);
%    \draw[tuh] (L) to (Z);
%    \draw[tuh] (L) to (X);
    \draw[huh] (X) edge (Z);
    \draw[tuh] (Z) to (Y);
    \node at (1,-0.75) {(b)};
  \end{scope}
  \begin{scope}[xshift=8cm,yshift=0cm]
    \node[ndout] (X) at (0,0) {$X$};
    \node[ndout] (Y) at (2,0) {$Y$};
    \node[ndout] (Z) at (1,1) {$Z$};
%    \node[ndlat] (L1) at (0,2) {$L_1$};
%    \node[ndlat] (L2) at (2,2) {$L_2$};
    \draw[tuh] (X) to (Y);
%    \draw[tuh] (L1) to (Z);
%    \draw[tuh] (L1) to (X);
    \draw[huh] (X) edge (Z);
%    \draw[tuh] (L2) to (Z);
%    \draw[tuh] (L2) to (Y);
    \draw[huh] (Z) edge (Y);
    \node at (1,-0.75) {(c)};
  \end{scope}
  \begin{scope}[xshift=0cm,yshift=-3cm]
    \node[ndout] (X) at (0,0) {$X$};
    \node[ndout] (Y) at (2,0) {$Y$};
    \node[ndout] (Z) at (1,1) {$Z$};
    \draw[tuh] (X) to (Y);
    \draw[tuh] (X) to (Z);
    \draw[tuh] (Z) to (Y);
    \node at (1,-0.75) {(d)};
  \end{scope}
  \begin{scope}[xshift=4cm,yshift=-3cm]
%      \draw[shadow] (-1,-2) rectangle (3,2);
    \node[ndout] (X) at (0,0) {$X$};
    \node[ndout] (Y) at (2,0) {$Y$};
    \node[ndout] (Z) at (1,1) {$Z$};
%    \node[ndlat] (L) at (1,-1) {$L$};
    \draw[tuh] (X) to (Y);
    \draw[tuh] (Z) to (X);
    \draw[tuh] (Z) to (Y);
%    \draw[tuh] (L) to (X);
%    \draw[tuh] (L) to (Y);
    \draw[huh,bend right] (X) edge (Y);
    \node at (1,-0.75) {(e)};
  \end{scope}
\end{tikzpicture}
\caption{\label{fig:simpson_paradox_causal_hypotheses}Different causal hypotheses (causal graphs of an L-CBN or SCM) for explaining the patient data.}
\end{figure}

In this exercise we will make use of causal reasoning with L-CBNs or SCMs to resolve Simpson's paradox.
Consider the following binary variables:
\begin{center}\begin{tabular}{llll}
  symbol & name & domain & meaning \\
  \hline
  $X$ & treatment & $\C{X} = \{0,1\}$ & 0=no drug, 1=drug \\
  $Y$ & outcome   & $\C{Y} = \{0,1\}$ & 0=died, 1=recovered \\
  $Z$ & sex       & $\C{Z} = \{0,1\}$ & 0=male, 1=female
\end{tabular}\end{center}
We will consider different possible causal hypotheses to explain the observed data.
In the following questions, you may assume that strict positivity holds, whenever convenient.
\begin{enumerate}[label=\alph*)]
  \item 
    Suppose that the data is generated by a causal model with causal graph depicted in Figure~\ref{fig:simpson_paradox_causal_hypotheses}(a).
    \begin{itemize}
      \item How is the Markov kernel
    $$P(Y=1 \given \doit(X=x))$$
    related to the observed distribution $P(X,Y,Z)$?
    If it is identifiable, derive an identity that expresses the former in terms of the latter and prove that this identity holds.
    If it is not identifiable, provide tight upper and lower bounds of the former in terms of the latter.
      \item Do the same for the Markov kernel:
    $$P(Y=1 \given Z=z,\doit(X=x)).$$
      \item Would you advise a patient to take the drug? Would your advice depend on the sex of the patient? 
         Motivate your answer using the expressions just derived and the contingency tables above.
    \end{itemize}
    \score{4}
    \answer{The first point can be ansered using the back-door criterion, or using the do-calculus.
    The back-door criterion immediately yields an expression, noting that $\{Z\}$ is the valid adjustment set.
    $$P(Y=1 \given \doit(X=x)) = \sum_{z \in \C{Z}} P(Y=1 \given Z=z, X=x) P(Z=z) \text{ a.s.}$$
    Alternatively one can apply the chain rule and the do-calculus with $Y \Perp I_X \given X, Z$ and $Z \Perp I_X$.
    The conditional probabilities $P(Y=1 \given Z=z, X=x)$ correspond with the recovery rates in the sex-specific tables. 
    This implies that a patient of unknown sex is better off by \emph{not} taking the drug.

    For the second point, the conditional back-door criterion (with empty adjustment set), or the do-calculus with $Y \Perp I_X \given X, Z$, gives:
    $$P(Y=1 \given Z=z,\doit(X=x)) = P(Y=1 \given Z=z,X=x) \text{ a.s.}$$
    Hence, patients of both sexes are better off by \emph{not} taking the drug.}
    \points{1 pt for the derivation of the causal effect, 1 pt for a well-motivated sensible corresponding advice.
    1 pt for the derivation of the conditional causal effect, 1 pt for a well-motivated sensible corresponding advice.}
  \item How would your answers for question a) change for the hypothesis in Figure~\ref{fig:simpson_paradox_causal_hypotheses}(b)?
    \score{1}
    \answer{They stay the same.}
  \item Answer the same questions as in a), but now for the causal hypothesis in Figure~\ref{fig:simpson_paradox_causal_hypotheses}(c).
    \score{4}
    \answer{Under this hypothesis, $\{Z\}$ is no longer a valid adjustment set, but $\emptyset$ is, and hence:
    $$P(Y=1 \given \doit(X=x)) = P(Y=1 \given X=x) \text{ a.s.}$$
    This also follows from the do-calculus with $Y \Perp I_X \given X$.
    These conditional probabilities correspond with the recovery rates in the aggregated table.
    Hence, for a patient of unknown sex the rational choice is to take the drug.

    There exists no valid adjustment set for the conditional treatment effect $P(Y=1 \given Z=z,\doit(X=x))$.
    Also the do-calculus will not help to identify this quantity (it is not identifiable from the data).
    One can derive lower and upper bounds on these probabilities, though, using the natural bounds.
    \begin{equation*}\begin{split}
      P(X=x,Y=1 \mid Z=z) 
      &\le P(Y=1 \mid Z=z, \doit(X=x)) \\
      &\le P(X=x,Y=1 \mid Z=z) + P_M(X\ne x|Z=z).
    \end{split}\end{equation*}

    \centerline{\begin{tabular}{llll}
      $x$ & $z$ & LB   & UB \\
      \hline
      0   & 0   & $700/4000 = 0.175$ & $3700/4000=0.925$ \\
      1   & 0   & $1800/4000 = 0.45$ & $2800/4000=0.70$ \\
      0   & 1   & $900/4000 = 0.225$ & $1900/4000=0.475$ \\
      1   & 1   & $200/4000 = 0.05$  & $3200/4000=0.80$ 
    \end{tabular}}

    Therefore, no sex-specific advice can be rationally concluded from these data.
    (Which is funny because we still can conclude that \emph{on average} it's best to take the drug).
    }
    \points{1 pt for the derivation of the causal effect, 1 pt for a well-motivated sensible corresponding advice.
    1 pt for stating bounds on the conditional causal effect, 1 pt for a well-motivated sensible corresponding advice.}
  \item The same questions for hypothesis Figure~\ref{fig:simpson_paradox_causal_hypotheses}(d).
    \score{4}
    \answer{The empty set is still valid for adjustment:
    $$P(Y=1 \given \doit(X=x)) = P(Y=1 \given X=x) \text{ a.s.}$$
    which can also be derived from $Y \Perp I_X \given X$.
    Hence, for a patient of unknown sex the rational choice is to take the drug.

    In this case, we can also estimate the conditional treatment effect:
    $$P(Y=1 \given Z=z,\doit(X=x)) = P(Y=1 \given Z=z,X=x) \text{ a.s.}$$
    However, this may be misleading, because $Z$ is not a ``pre-treatment'' variable (but is caused by treatment under this hypothesis). 
    $Z$ now corresponds to sex \emph{after} treatment, which may differ from sex before treatment
    (perhaps the drug has the side effect of changing sex in some patients). 

%    Na\"ively interpreting this expression counterfactually seems to tell us that for those patients that are male or female after the treatment, they would have been better off without the drug.
%    Yet, we would advice a patient of unknown sex to take the drug.
%    That doesn't make sense!
%    But it's indeed not the correct interpretation.
%    Counterfactually, patients that are male or female after treatment, would have been better of without treatment, iff:
%      P(Y=1 | Z=0, do(X=1)) > P(Y'=1 | Z=0, do(X=1), do(X'=0))  AND  P(Y=1 | Z=1, do(X=1)) > P(Y'=1 | Z=1, do(X=1), do(X'=0))
%    One can prove easily (just the chain rule, no other assumptions needed) that this implies:
%      P(Y=1 | do(X=1)) > P(Y'=1 | do(X'=0))
%    So indeed the na\"ive interpretation made no sense.
%
%    So my conclusion is that the quantity $P(Y = 1 \given Z=z,\doit(X=x))$ seems to carry no clear decision-theoretic meaning.

    So what looks mathematically like a `conditional treatment effect' does not tell anything useful for deciding on optimal actions under this hypothesis.
    If we wish to give sex-specific advice on whether someone should take the drug, we should condition on sex \emph{before} treatment, which is not available in the data (at least under this causal hypothesis, where $Z$ is an effect of $X$).

    So what we can conclude from the data under this hypothesis is only that recovery will be higher if we treat the entire population than if we treat noone.}
    \points{1 pt for the derivation of the causal effect, 1 pt for a well-motivated sensible corresponding advice.
    1 pt for stating bounds on the conditional causal effect, 1 pt for a well-motivated sensible corresponding advice.}
  \item The same question for hypothesis Figure~\ref{fig:simpson_paradox_causal_hypotheses}(e).
    \score{4}
    \answer{Under this hypothesis, both quantities cannot be identified (at best, they can be bounded).
    
    Using the natural bounds, we obtain lower and upper bounds on $P(Y=1 \mid \doit(X=x))$: 
    \begin{equation*}\begin{split}
      P(X=x,Y=1) 
      &\le P(Y=1 \mid \doit(X=x)) \\
      &\le P(X=x,Y=1) + P_M(X\ne x).
    \end{split}\end{equation*}

    \centerline{\begin{tabular}{llll}
      $x$ & LB   & UB \\
      \hline
      0   & $1600/8000 = 0.20$ & $5600/8000=0.7$ \\
      1   & $2000/8000 = 0.25$ & $6000/4000=0.75$ \\
    \end{tabular}}

    The lower and upper bounds for $P(Y=1 \mid \doit(X=x), Z=z)$ are the same as for hypothesis (c).

    The bounds do not allow us to draw any conclusion on whether it's better to take the drug or not (both unconditional on sex, and conditional on sex).
}
    \points{1 pt for stating bounds on the causal effect, 1 pt for a well-motivated sensible corresponding advice.
    1 pt for stating bounds on the conditional causal effect, 1 pt for a well-motivated sensible corresponding advice.}
  \item Would you consider all hypotheses (a)-(e) equally likely, or would you reject some of these \emph{a priori} based on common sense background knowledge about causal relations? Motivate your answer.
  \score{2}
  \answer{A reasonable assumption seems to be that treatment $X$ does not cause sex $Z$, allowing to reject (d).
  If we rule out selection bias in the past (the male/female ratio in the data is 50\%, so the data do not contain evidence that there has been selection bias), then it also seems unlikely that there are unobserved common causes of $Z$ and $X$, and of $Z$ and $Y$.
  This would rule out hypotheses (b) and (c). 
  But the validity of that assumption is debatable (and untestable from the available data).
  }
  \points{If a student interprets a bidirected edge $A \huh B$ as that $A$ and $B$ cause each other (a cycle), zero points (that is so wrong).}
  \item A law suit is filed against the pharmaceutical company that developed the drug, because several male patients who took the drug died quickly afterwards.
    The plaintiff claims that she inferred from the data that the probability that these patients would have recovered (and hence still be alive) had they not been treated, is 70\%.
    The court asks you for your expert advice in the matter. 
    What would be your response?
  \score{2}
  \answer{Such counterfactual quantities are generically unidentifiable from observed data, without making very strong modeling assumptions.
    As reliable models probably don't exist (given that the disease is new and the drug is experimental), the inferred probability has to be taken with a large grain of salt.
    At best, the data may lead to a lower and an upper bound on the counterfactual probability in the claim, instead of a point estimate as provided by the data scientist.}
    \points{The answer ``the claim is not founded on a causal model'' is worth 1 point.}
  \item Assume that treatment $X$ does not cause sex $Z$. 
    \score{3}
    Prove that Simpson's paradox cannot occur if the data come from a proper randomized controlled trial, in which treatment $X$ is completely randomized.
    \answer{We model $X$ as an exogenous input variable.
    We may assume the following CDMG.
    \begin{center}\begin{tikzpicture}
    \begin{scope}[xshift=8cm,yshift=-3cm]
  %      \draw[shadow] (-1,-2) rectangle (3,2);
      \node[ndint] (X) at (0,0) {$X$};
      \node[ndout] (Y) at (2,0) {$Y$};
      \node[ndout] (Z) at (1,1) {$Z$};
  %    \node[ndlat] (L) at (1,-1) {$L$};
      \draw[tuh] (X) to (Y);
  %    \draw[tuh] (Z) to (X);
      \draw[tuh] (Z) to (Y);
  %    \draw[tuh] (L) to (X);
  %    \draw[tuh] (L) to (Y);
      \draw[huh,bend left] (Z) edge (Y);
      \draw[tuh,bend left] (Y) edge (Z);
    \end{scope}\end{tikzpicture}\end{center}
%Suppose treatment is beneficial for both sexes.
%$$P(Y_1 = 1 \given Z=z) > P(Y_0 = 1 \given Z=z)$$
%for all $z$ implies
%$$P(Y_1 = 1) = \sum_{z} P(Y_1 = 1 \given Z=z) P(Z=z)
%             > \sum_{z} P(Y_0 = 1 \given Z=z) P(Z=z)
%             = P(Y_0 = 1)$$
%Hence, treatment is beneficial overall.
%
Now
$$P(Y = 1 \given Z=z, \doit(X=1)) > P(Y = 1 \given Z=z, \doit(X=0)) \text{ a.s.}$$
for all $z$ implies
  \begin{equation*}\begin{split}
  P(Y = 1 \given \doit(X=1)) 
    &= \sum_{z} P(Y = 1 \given Z=z, \doit(X=1)) P(Z=z \given \doit(X=1)) \\
    &= \sum_{z} P(Y = 1 \given Z=z, \doit(X=1)) P(Z=z \given \doit(X=0)) \\
    &> \sum_{z} P(Y = 1 \given Z=z, \doit(X=0)) P(Z=z \given \doit(X=0)) \\
    &= P(Y = 1 \given \doit(X=0)) \text{ a.s.}
  \end{split}\end{equation*}
Here we first used the chain rule, and then the fact that $X$ does not cause $Z$ 
(just the definition of rung-2 causation).

If one interprets the assumption ``$X$ does not cause $Z$'' graphically, then
the edge $Z \hut Y$ should be absent, and $P(Z=z \given \doit(X=1)) = P(Z=z \given \doit(X=0))$ follows from the Markov property and $Z \Perp X$.

Hence, the weird reversal of which treatment is best under additional conditioning on sex cannot occur.
%
%On the other hand,
%$$P(Y = 1 \given X=1, Z=z) > P(Y = 1 \given X=0, Z=z)$$
%for all $z$ does \emph{not} imply
%  \begin{equation*}\begin{split}
%  P(Y = 1 \given X=1) 
%    &= \sum_{z} P(Y = 1 \given X=1, Z=z) P(Z=z \given X=1) \\
%    &\ne \sum_{z} P(Y = 1 \given X=1, Z=z) P(Z=z \given X=0) \\
%    &> \sum_{z} P(Y = 1 \given X=0, Z=z) P(Z=z \given X=0) \\
%    &= P(Y = 1 \given X=1)
%  \end{split}\end{equation*}
  }
  \points{A similar answer but where $X$ is modeled as an endogenous variable (without confounding or causes) is also correct.}
\end{enumerate}
